---
title: "Interactive Web Applications in R with Shiny"
subtitle: "Ann Arbor R User Group"
author: "Clayton Yochum"
date: "June 3, 2015"
output:
  ioslides_presentation:
    widescreen: yes
    incremental: yes
---

## Who am I? {.build}

- Co-organizer of this Meetup group
- Business Intelligence Analyst at Barracuda Networks
- R user for almost 3 years, Shiny user for ~2 years


## What is Shiny? {.build}

- It's a web framework for R!
- Primarily created & maintained by the RStudio Group
- open source (well, mostly)!


## How does it work? {.build}

- You write all your code in R
- Use a reactive programming paradigm to write both server and client code
- Shiny abstracts away the HTML, CSS, and Javascript


## Examples {.build}

- [almost cool](http://shiny.rstudio.com/gallery/single-file-shiny-app.html)
- [a little cooler](http://shiny.rstudio.com/gallery/kmeans-example.html)
- [even cooler than that](http://shiny.rstudio.com/gallery/movie-explorer.html)


# My First App


## Hello, World! {.build}

- all you need is is the function `shinyApp`, which takes arguments `ui` and `server`
```{r, eval=FALSE}
install.packages(shiny)
library(shiny)

shinyApp(ui=fluidPage(), server=function(input, output) {})
shinyApp(ui=fluidPage("Hello, World!"), server=function(input, output) {})
shinyApp(ui=fluidPage(h1("Hello, World!")), server=function(input, output) {})
```
- yes, it really is that easy!
- open it in a browser
- check the console


## What's really going on here? {.build}

- let's run some of those ui functions on their own: <br>

```{r, eval=FALSE}
fluidPage()
fluidPage("Hello, World!")
fluidPage(h1("Hello, World!"))
```

- it's just HTML!
- the server function takes two inputs and just doesnt do anything: <br>

```{r, eval=FALSE}
server <- function(input, output) {}
server("foo", "bar")
.Last.value
```

- check out the source of our "Hello, World!" program in the browser


## Running from a file {.build}

- rather than stuffing the whole application in a command, let's make a file `app.R`:

```{r, eval=FALSE}
# app.R

ui <- fluidPage(h1("Hello, World!"))

server <- function(input, output) {}

shinyApp(ui=ui, server=server)
```
- run it with `runApp()`
- send it straight to the browser with `runApp(launch.browser=TRUE)`


## Where's the data? {.build .smaller}

- Let's generate some (boring) random data and display it:

```{r, eval=FALSE}
# app.R

ui <- fluidPage(
  h1("Hello, World!"),
  tableOutput("data")
)

server <- function(input, output) {
  output$data <- renderTable(
    data.frame(letter = letters[1:10])
  )
}

shinyApp(ui=ui, server=server)
```
- notice how we assign to `output$data` in `server`, but use the string `"data"` in `ui`


## Adding an input {.build .smaller}

- use `selectInput()`:
```{r, eval=FALSE}
# app.R

ui <- fluidPage(
  h1("Hello, World!"),
  sliderInput("slider", "choose a length for the dataset:", 1, 26, 10),
  tableOutput("data")
)

server <- function(input, output) {
  output$data <- renderTable(
    data.frame(letter = letters[1:10])
  )
}

shinyApp(ui=ui, server=server)
```
- run it!


## Hey, nothing happened! {.build .smaller}

- well of couse not, we never changed the server function:
```{r, eval=FALSE}
# app.R

ui <- fluidPage(
  h1("Hello, World!"),
  sliderInput("slider", "choose a length for the dataset:", 1, 26, 10),
  tableOutput("data")
)

server <- function(input, output) {
  output$data <- renderTable(
    data.frame(letter = letters[1:input$slider])
  )
}

shinyApp(ui=ui, server=server)
```
- now it works!


## Better Data Handling {.build .smaller}

- let's generate a more complex dataset, and sample some rows to display:
```{r, eval=FALSE}
# app.R

ui <- fluidPage(
  h1("Hello, World!"),
  sliderInput("slider", "choose a length for the dataset:", 1, 26, 10),
  tableOutput("data")
)

server <- function(input, output) {
  output$data <- renderTable({
    df <- data.frame(a = runif(10000000), b = rnorm(10000000))
    rows <- sample(nrow(df), input$slider)
    df[rows, ]
  })
}

shinyApp(ui=ui, server=server)
```
- we needed the curly braces in `renderTable` because our expression is more than one line
- why is the response time so crappy?


## Better Data Handling {.build .smaller}
- we were generating the dataset from scratch each time we moved the slider
- we can avoid that by generating `df` _before_ calling `renderTable`:
```{r, eval=FALSE}
# app.R

ui <- fluidPage(
  h1("Hello, World!"),
  sliderInput("slider", "choose a length for the dataset:", 1, 26, 10),
  tableOutput("data")
)

server <- function(input, output) {
  df <- data.frame(a = runif(10000000), b = rnorm(10000000))
  output$data <- renderTable({
    rows <- sample(nrow(df), input$slider)
    df[rows, ]
  })
}

shinyApp(ui=ui, server=server)
```
- so much faster! We only generated `df` *once*, when we first _connected_ to the app
- only the `renderTable` call runs when `input$slider` changes; that's reactivity!


# Best Practices


## Separating the Front and Back Ends {.build .smaller}

- put the front end in a file `ui.R`, and the backend in `server.R`:
```{r, eval=FALSE}
# ui.R

fluidPage(
  h1("Hello, World!"),
  sliderInput("slider", "choose a length for the dataset:", 1, 26, 10),
  tableOutput("data")
)
```
```{r, eval=FALSE}
# server.R

function(input, output) {
  df <- data.frame(a = runif(10000000), b = rnorm(10000000))
  output$data <- renderTable({
    rows <- sample(nrow(df), input$slider)
    df[rows, ]
  })
}
```
- remove the old app.R (`file.remove("app.R")`) and run it the same way as before
- we don't need to assign the `fluidPage`/function to variables anymore


## Global Variables {.build .smaller}

- try refreshing the page; see that delay?
- currently, our `df` is created when the function in `server.R` runs, which is when a client (browser) connects
- we can make the variable _global_ by breaking it out of the function inside `server.R`:
```{r, eval=FALSE}
# server.R

df <- data.frame(a = runif(10000000), b = rnorm(10000000))

function(input, output) {
  output$data <- renderTable({
    rows <- sample(nrow(df), input$slider)
    df[rows, ]
  })
}
```
- refresh; faster, right?
- now `df` is generated once when we call `runApp`, instead of each time someone connects


## global.R {.build}

- we can also break out global operations into their own file, `global.R`:
```{r, eval=FALSE}
# global.R

df <- data.frame(a = runif(10000000), b = rnorm(10000000))
```
- don't forget to remove that line from `server.R`
- run it; behaviour hasn't changed
- why do this?


## functions.R {.build .smaller}

- at Barracuda, we like to remove as much code from `server.R` and `global.R` as possible
- this makes the logic easier to read, particularly when your reactive model becomes complex
- put all the workhorse code in _functions_ inside a file like `functions.R`:
```{r, eval=FALSE}
# functions.R

makeData <- function() {
  data.frame(a = runif(10000000), b = rnorm(10000000))
}

makeTable <- function(df, n) {
  rows <- sample(nrow(df), n)
  df[rows, ]
}
```
- you can call this file whatever you like, as we have to source it in ourselves

## functions.R {.build .smaller}

- source `functions.R` in `global.R`, and replace the old code to make `df`:
```{r, eval=FALSE}
# global.R

source("functions.R")
df <- makeDate()
```
- and to generate the table in `server.R`:
```{r, eval=FALSE}
function(input, output) {
  output$data <- renderTable(makeTable(df, input$slider))
}
```
- can you put the `renderTable` call in `makeTable`? try it out


# Deployment


## Deploying on the local network {.build}

- so far, the applications we've ran have only been available to our local machine
- to open it up to your local network, add a `host` arg to `runApp`:
```{r, eval=FALSE}
runApp(launch.browser=TRUE, host="0.0.0.0")
```
- now change the URL in your browser to use your IP address instead of `127.0.0.1`
- try hitting that URL with a browser on a different device on the same local network
- great for getting feedback from coworkers & managers during development


## Deploying online with shinyapps {.build}

- a hosted service from RStudio
- uses a special `shinyapps` package to deploy right from the R console
- free and paid tiers available (more applications and run-hours)
- see more at their [website](www.shinyapps.io)


## Deploying on your own server {.build}

- ShinyServer is a free, open source hosting package from RStudio
- intended for running on your own linux server
- Pro (closed source) version adds a ton of things like LDAP integration, management console, and load balancing
- [Website](http://www.rstudio.com/products/shiny/shiny-server/)

## What else can I do with shiny? {.build}

- better layout (`column`/`fluidRow`, `pageWithSidebar`)
- more reactivity (chains of dependency, complex behaviours)
- more inputs (`selectInput`, `checkboxInput`, `checkboxGroupInput`, `radioButtons`)
- hide or change inputs based on other inputs/outputs (including server-side input generation)
- enhance shiny with packages like:
  + `rCharts` & `leaflet` (`D3.js` visualizations)
  + `datatables` (interactive tabular output)
  + `shinydashboard` (ui tools for making attractive dashboards)
- add shiny widgets to an RMarkdown file
- hack at the HTML/CSS/Javascript


## Pros & Cons {.build}

Pros:

- lets you create and share live, interactive web content
- no knowledge of HTML/CSS/JavaScript needed
- great extensions available
- very good documentation and community support

Cons:

- limited options for free deployment
- harsh learning curve (esp. for complex reactivity)


## Resources

- [Official page](http://shiny.rstudio.com/) (lots of articles/tutorials)
- [Github](https://github.com/rstudio/shiny)
- [Webinars](http://www.rstudio.com/resources/webinars/)
- [StackOverflow tag](https://stackoverflow.com/questions/tagged/shiny)
- [Google Groups](https://groups.google.com/forum/#!forum/shiny-discuss) (ugh)
- [Official Examples](http://shiny.rstudio.com/gallery/)
- [Show Me Shiny](http://www.showmeshiny.com/)


# Time to add a plot?


# Questions? Comments?
