---
title: "Introduction to ggvis"
subtitle: "Ann Arbor R User Group"
author: "Dave Childers"
date: "October 15, 2015"
output: ioslides_presentation
runtime: shiny
---

<!--
incremental: no
smaller: true

-->

## Intro

Dave Childers

- Data Scientist with [Powerley](http://www.powerley.com/)
- Before that: Data Science Consulting
- Before that: CSCAR at UM

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(shiny)
library(knitr)
library(ggplot2)
library(ggvis)
library(dplyr)
library(scales)
library(DiagrammeR)

cocaine <- tbl_df(cocaine)
```

## Example Data Set | All Cocaine Seizures in 2007
```{r, echo = FALSE, results="asis"}
cocaine %>% slice(1:5) %>% as.data.frame %>% knitr::kable(.)
```

<!--

## Motivation

```{r, echo=FALSE}
cocaine %>%
  mutate(ppg = price / weight) %>%
  group_by(state) %>%
  summarise(
    mean_potency = mean(potency), 
    mean_ppg = mean(ppg),
    n = n()
    ) %>%
  filter(n >= 10) %>%
  mutate(id = state) %>% 
  ggvis(~mean_potency, ~mean_ppg, key := ~id) %>%
  layer_points %>%
  add_tooltip(function(x) x$id) %>%
  add_axis("y", title = "Average Cost per Gram") %>%
  add_axis("x", title = "Average Potency")
```

-->

## Motivation

```{r, echo=FALSE}
cocaine %>%
  slice(1:100) %>%
  ggvis(~weight , ~price) %>%
  layer_points(
    size := input_slider(min = 50, max = 500, value = 50, step = 50), 
    opacity := 0.3,
    fill := "steelblue"
    ) %>%
  layer_model_predictions(
    model = input_radiobuttons(
      choices = c("Linear" = "lm", "LOESS" = "loess"),
      selected = "loess",
      label = "Model Type"
    ),
    formula = price ~ weight
  ) #%>%
  #bind_shiny("p")

#ggvisOutput("p")
```

## Motivation

- Graphics moving to web
- Interactivity
- Fast Exploratory Data Analysis

## Outline

1. Dependencies
2. Syntax
3. Exploratory Data Analysis
4. Interactivity

## ggvis dependencies

```{r, echo=FALSE}
grViz("
      
digraph ggvis {

# add node statements
node [shape = circle
fonrname = Helvetica
penwidth = 2.0]

ggvis; dplyr; shiny; vega; D3; gg; magrittr

# add edge statements
edge[arrowhead = diamond]
dplyr->ggvis; shiny->ggvis; vega->ggvis;
D3->ggvis; gg->ggvis; magrittr->ggvis

# add a graph statement
graph [nodesep = 0.1]
}      
")
```

<!--

## dplyr | grammar of data manipulation

```{r}
filter(cocaine, weight <= 10) # dplyr version of base::subset
mutate(cocaine, log_weight = log(weight)) # dplyr version of base::transform
```

[dplyr introduction](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)

-->

## pipe operator | read code like a book

```
# easy to read/write
cocaine %>%
  mutate(log_weight = log(weight)) %>%
  filter(weight <= 10)

# hard to read/write
filter(mutate(cocaine, log_weight = log(weight)), weight <= 10)
```

[pipe introduction](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)

# Syntax

## ggplot

<!--

```
### <b>
ggplot(cocaine, aes(x = weight, y = price)) +
  geom_point()
### </b>
```
-->

```{r}
ggplot(cocaine, aes(x = weight, y = price)) +
  geom_point()
```

## ggvis

```{r}
ggvis(cocaine, x = ~weight, y = ~price) %>%
  layer_points
```

## ggplot v ggvis | Syntax Comparison

```
ggplot(cocaine, aes(x = weight, y = price)) +
  geom_point()

ggvis(cocaine, x = ~weight, y = ~price) %>%
  layer_points
```

__Similarities__

- Define graphic by composing *small* functions
- Mapping variables to visual properties
- Property Inheritance

## ggplot v ggvis | Syntax Comparison

```
ggplot(cocaine, aes(x = weight, y = price)) +
  geom_point()

ggvis(cocaine, x = ~weight, y = ~price) %>%
  layer_points
```

__Differences (ggplot -> ggvis)__

  1. plus -> pipe
  2. layer -> geom
  3. point -> points
  4. aes() -> ~

## Table of Geoms/Layers

```{r, results="asis", echo=FALSE}
read.csv("./config/geoms_layers.csv") %>% knitr::kable(.)
```

## No Layers

```{r, results="asis", echo=FALSE}
read.csv("./config/no_layers.csv") %>% knitr::kable(.)
```

## Setting Constants, Mapping Variables

```
# mapping state -> color
ggplot(cocaine, aes(x = weight, y = price)) +
  geom_point(aes(color = state))

# setting color to a constant value
ggplot(cocaine, aes(x = weight, y = price)) +
  geom_point(color = "orange")
```

```{r, echo=FALSE}
c2 <- cocaine %>% filter(state %in% c("FL", "IN", "NY"))
```

## Variables/Constants Mapping/Setting
```{r, results="asis", echo=FALSE}
read.csv("./config/where_how.csv") %>% 
  setNames(., c("", "Variable", "Constant")) %>%
  knitr::kable(.)
```

<!--

## ggplot Mapping

```{r}
ggplot(c2, aes(x = weight, y = price)) +
  geom_point(aes(color = state))
```

## ggplot Setting

```{r}
ggplot(c2, aes(x = weight, y = price)) +
  geom_point(color = "orange")
```

-->

# Exploratory Data Analysis | Cocaine Seizures

<!--

## Histograms

```{r}
ggvis(cocaine, x = ~price) %>% layer_histograms
```

-->

## Histograms

```{r}
ggvis(cocaine, x = ~price) %>% 
  layer_histograms(width = 100, center = 50)
```

## Densities

```{r}
ggvis(cocaine, x = ~price) %>% layer_densities %>% 
  add_axis("y", title_offset = 50)
```

<!--

## Densities Options

```{r}
ggvis(cocaine, x = ~price) %>% layer_densities(adjust = 0.3, area = FALSE)
```

## Counts by State

```{r}
cocaine %>% count(state, sort = TRUE)
```

-->

## Bar Plot

```{r}
ggvis(cocaine, x = ~state) %>% layer_bars
```

<!--

## Ordered Bar Plot

```{r}
cocaine %>% count(state, sort = TRUE) %>%
  ggvis(x = ~reorder(state, -n), y = ~n) %>%
  layer_bars
```

-->

## Bar Plot #2

```{r, echo=FALSE}
cocaine %>% count(state, sort = TRUE) %>%
  ggvis(x = ~reorder(state, -n), y = ~n) %>%
  filter(n >= 100) %>%
  add_axis("x", title = "State") %>%
  layer_bars %>%
  layer_text(text := ~n, fontSize := 20) 
```

What happens if we say `text = ~n` ?


## Bar Plot Code

```
cocaine %>% count(state, sort = TRUE) %>%
  ggvis(x = ~reorder(state, -n), y = ~n) %>%
  filter(n >= 100) %>%
  add_axis("x", title = "State") %>%
  layer_bars %>%
  layer_text(text := ~n, fontSize := 20) 
  # what happens if we say text = ~n ?
```

## Title? 

```{r, echo = FALSE}
cocaine %>% count(state, sort = TRUE) %>%
  ggvis(x = ~reorder(state, -n), y = ~n) %>%
  filter(n >= 100) %>%
  layer_bars %>%
  layer_text(text := ~n) %>%
  add_axis("x", title = "State") %>%
  add_axis(
    "x", 
    title = "2007 Cocaine Seizures by State", 
    orient = "top",
    ticks = 0,
    properties = axis_props(
      axis = list(stroke = "white"),
      labels = list(fontSize = 0)
    )
    )
```

## Title? | A Hack

```
cocaine %>% count(state, sort = TRUE) %>%
  ggvis(x = ~reorder(state, -n), y = ~n) %>%
  filter(n >= 100) %>%
  layer_bars %>%
  layer_text(text := ~n) %>%
  add_axis("x", title = "State") %>%
  add_axis(
    "x", 
    title = "2007 Cocaine Seizures by State", 
    orient = "top",
    ticks = 0,
    properties = axis_props(
      axis = list(stroke = "white"),
      labels = list(fontSize = 0)
    )
    )
```

## Not (Yet) Implemented

- ```ggtitle()```
- ```coord_flip()```
- ```themes```
- Faceting

## Smooths

```
%>% layer_smooths
```

```{r, echo=FALSE}
ggvis(cocaine, x = ~weight, y = ~price) %>%
  layer_points(opacity := 0.2, fill := "steelblue") %>%
  layer_smooths
```

## Linear Model

```
%>% layer_model_predictions(model = "lm", formula = price ~ weight)
```

```{r, echo=FALSE}
ggvis(cocaine, x = ~weight, y = ~price) %>%
  layer_points(opacity := 0.2, fill := "steelblue") %>%
  layer_model_predictions(model = "lm", formula = price ~ weight)
```

<!--

## Log-log model

```{r log_log, eval=FALSE}
ggvis(cocaine, ~weight, ~price) %>%
  mutate(
    log_weight = log(weight), 
    log_price = log(price)
    ) %>%
  layer_points(x = ~log_weight, y = ~log_price, opacity := 0.2, fill := "steelblue") %>%
  layer_model_predictions(
    model = "lm", 
    formula = log_price ~ log_weight
    )
```

# Log-log model

```{r log_log2, echo=FALSE}
ggvis(cocaine, ~weight, ~price) %>%
  mutate(
    log_weight = log(weight), 
    log_price = log(price)
    ) %>%
  layer_points(x = ~log_weight, y = ~log_price, opacity := 0.2, fill := "steelblue") %>%
  layer_model_predictions(
    model = "lm", 
    formula = log_price ~ log_weight
    )
```

## Filter

```{r}
cocaine %>% count(state, sort = TRUE) %>%
  ggvis(x = ~reorder(state, -n), y = ~n) %>%
  filter(n >= 100) %>%
  add_axis("x", title = "State")
```

-->

## Smooths by state

```{r, echo=FALSE}
ggvis(cocaine, ~weight, ~price, fill = ~state, stroke = ~state) %>%
  filter(state %in% c("FL", "IN", "NY")) %>%
  mutate(log_weight = log(weight), log_price = log(price)) %>%
  layer_points(~log_weight, ~log_price, opacity := 0.2) %>%
  auto_group() %>%
  layer_model_predictions(
    model = "lm", 
    formula = log_price ~ log_weight
    ) %>%
  add_axis("x", title = "Log Price") %>%
  add_axis("y", title = "Log Weight")
```

## Smooths by state | Code

```
ggvis(cocaine, ~weight, ~price, fill = ~state, stroke = ~state) %>%
  filter(state %in% c("FL", "IN", "NY")) %>%
  mutate(log_weight = log(weight), log_price = log(price)) %>%
  layer_points(~log_weight, ~log_price, opacity := 0.2) %>%
  auto_group() %>%
  layer_model_predictions(
    model = "lm", 
    formula = log_price ~ log_weight
  ) %>%
  add_axis("x", title = "Log Price") %>%
  add_axis("y", title = "Log Weight")
```

## dplyr methods on ggvis objects

```{r}
methods(class = "ggvis")
```

<!--

```{r, results="asis", echo=FALSE}
read.csv("./config/no_layers.csv") %>% knitr::kable(.)
```

-->

## ggvis scales options 

```{r, results="asis", echo=FALSE}
read.csv("config/ggvis_scales.csv") %>% #setNames("option") %>%
  knitr::kable(.)
```

# Interactivity

## Control Size and Model

```{r, echo=FALSE}
cocaine %>%
  slice(1:100) %>%
  ggvis(~weight , ~price) %>%
  layer_points(
    size := input_slider(min = 50, max = 500, value = 50, step = 50), 
    opacity := 0.3,
    fill := "steelblue"
    ) %>%
  layer_model_predictions(
    model = input_radiobuttons(
      choices = c("Linear" = "lm", "LOESS" = "loess"),
      selected = "loess",
      label = "Model Type"
    ),
    formula = price ~ weight
  )
```

## Interactive Control Commands

```
%>% layer_points(
    size := input_slider(min = 50, max = 500, value = 50, step = 50)
)
```

<!--
%>% layer_model_predictions(
    model = input_radiobuttons(
      choices = c("Linear" = "lm", "LOESS" = "loess"),
      selected = "loess",
      label = "Model Type"
    )
-->

## Interactive Controls: Shiny & ggvis

```{r, echo=FALSE, results="asis"}
read.csv("config/interactive_controls.csv") %>% knitr::kable(.)
```

<!--

## Interactivity Possibilities

- arguments to transforms and layers

```layer_smooths(span = input_slider(0, 1))```

```layer_points(size := input_slider(1, 10))```

-->

## Tooltip
```{r, echo=FALSE}
cocaine %>%
  mutate(ppg = price / weight) %>%
  group_by(state) %>%
  summarise(
    mean_potency = mean(potency), 
    mean_ppg = mean(ppg),
    n = n()
    ) %>%
  filter(n >= 10) %>%
  mutate(id = state) %>%
  ggvis(~mean_potency, ~mean_ppg, key := ~id) %>%
  layer_points %>%
  add_tooltip(function(x) x$id) %>%
  add_axis("y", title = "Mean PPG") %>%
  add_axis("x", title = "Mean Potency")
```

## Tooltip Commands

```
%>% add_tooltip(function(x) x$id) 
```

## ggvis interactivity | Limitations

- cannot switch between data sets

- cannot add/remove layers

- need Shiny for full interactivity

## add ggvis to Shiny

```{r bind_shiny, results="asis", echo=FALSE}
data.frame(
  type = c("plot", "ggvis"),
  server.R = c("renderPlot", "bind_shiny"),
  ui.R = c("plotOutput", "ggvisOutput")
) %>% knitr::kable(.)
```

## Why ggvis?

- More **interactive** than ggplot 

- Faster **exploratory** analysis than shiny

- Be cautious about using in production

## Stackoverflow Tags

```{r, echo=FALSE}
tags <- read.csv("./config/so_tags.csv")
```

```{r, echo=FALSE}
ggvis(tags, x = ~tag, y = ~count) %>% 
  mutate(tag = reorder(tag, -count)) %>%
  layer_bars(fill = ~tag) %>%
  layer_text(text := ~ count, align := "left", fontSize := 30, dx := 60) %>%
  add_axis("y", title = "") %>%
  add_axis("x", title = "",
           properties = axis_props(
             ticks = list(stroke = "red"),
             labels = list(
               fill = "steelblue",
               angle = 45,
               fontSize = 16,
               baseline = "middle",
               dx = 3,
               dy = 10
             )
           )
  )
```

<!--

## Previous Code

```{r, eval=FALSE}
ggvis(tags, x = ~tag, y = ~count) %>% 
  mutate(tag = reorder(tag, -count)) %>%
  layer_bars(fill = ~tag) %>%
  layer_text(text := ~ count, align := "left", fontSize := 30, dx := 60) %>%
  add_axis("y", title = "") %>%
  add_axis("x", title = "",
           properties = axis_props(
             ticks = list(stroke = "red"),
             labels = list(
               fill = "steelblue",
               angle = 45,
               fontSize = 16,
               baseline = "middle",
               dx = 3,
               dy = 10
             )
           )
  )
```

-->

## Thanks! | Resources

- [ggvis webpage](http://ggvis.rstudio.com/)
- [Stackoverflow](http://stackoverflow.com/)
- [RStudio Blog](http://blog.rstudio.org/)

### Contact

- Dave Childers
- childers.dave@gmail.com
