---
title: "Introduction to INLA"
author: "Ellis Valentiner"
date: "October 24, 2014"
output:
  html_document:
    css: ~/Dropbox/HomePage/Styles/markdown-css-themes-gh-pages/markdown5.css
    theme: spacelab
---

```{r Setup, results='hide', warning=FALSE, message=FALSE}
rm(list = ls(all.names = TRUE))

library(INLA)
library(R2BayesX)
library(RColorBrewer)
library(stargazer)
library(ggplot2)
```

### Overview of Multilevel Modeling

Extension of regression for data with group structures, where coefficients can vary by group.

* Nested data
* Repeated measures
* Time-series cross sections
* Other non-nested structures

Some examples of data with group structures:

- Children within classrooms
- Patients within a clinical treatment
- Repeated measures for the same individual

Some advantages:

* accounts for variation at multiple levels when estimating group-level coefficients
* model variation among subject-level coefficients
* estimate coefficients for specific groups

### Bayesian Models

Considers parameters as random variables that are characterized by a prior distribution. This prior distribution is combined with the traditional likelihood to obtain the posterior distribution of the parameter of interest on which the statistical inference is based.

### Installing INLA

http://www.r-inla.org/download

```{r, eval=FALSE}
source("http://www.math.ntnu.no/inla/givemeINLA.R") 
```

INLA **requires** `sp` and suggests `numDeriv`, `Rgraphviz` (from Bioconductor), `fields`, `rgl`, `mvtnorm`, `multicore`, and `pixmap`

### Latent Models

Model | Name 
-----------------------------|------
Independent random variables | `iid`
Linear | `linear`
Constrained linear | `clienar`
Random walk of order 1 | `rw1`
Random walk of order 2 | `rw2`
Continuous random walk of order 2 | `crw2`
Model for seasonal variation | `seasonal`
Model for spatial effect | `besag`
Model for spatial effect | `besagproper`

### Example: Sleep Study

```{r}
#for lmer() and sleepstudy
library(lme4)

system.time(fm1 <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy))
system.time(fm2 <- inla(Reaction ~ Days + f(Subject), data = sleepstudy))

summary(fm1)
summary(fm2)
```

### Example: Munich

Semiparametric regression model with Gaussian data and additional spatial effect. Example  4.2.2 from "Gaussian Markov Random Fields: Theory and Applications"  by H. Rue and L. Held

```{r Munich Example}
#load the data
data(Munich)

## Note that here we what to have an estimator of the effect of `year`
## also the for years where we have no observation, therefore we give a
## vector with all possible values assumed by the covariate `year`, that
## is seq(1918,2001)

formula <- rent ~ f(location, model="besag", graph = "Munich/munich.graph",
                      initial = 1, param = c(1, 0.01)) +
                      f(year, model = "rw2", values = seq(1918, 2001),
                        param = c(1, 0.01)) +
                      f(floor.size, model = "rw2", param=c(1, 0.01)) +
                      Gute.Wohnlage + Beste.Wohnlage + Keine.Wwv + Keine.Zh +
                      Kein.Badkach  + Besond.Bad + Gehobene.Kueche +
                      zim1 + zim2 + zim3 + zim4 + zim5 + zim6 -1

mod <- inla(formula, data = Munich, verbose = TRUE,
                  control.family = list(initial = -1),
                  control.predictor = list(initial = 12),
                  control.inla = list(h = 1e-4))
```

```{r Print fixed effects, results='asis'}
fixed.effects <- mod$summary.fixed
colnames(fixed.effects) <- c('Mean', 'SD', 'Q0.025', 'Median', 'Q0.975', 'Mode', 'KLD')
stargazer(fixed.effects, type = 'html', title = 'Fixed effects')
```

```{r Print hyperparameters, results='asis'}
hyperparameters <- mod$summary.hyperpar
colnames(hyperparameters) <- c('Mean', 'SD', 'Q0.025', 'Median', 'Q0.975', 'Mode')
stargazer(hyperparameters, type = 'html', title = 'Hyperparameters')
```

### Example: Zambia

Semiparametric regression on Gaussian data with additional spatial effect. Example taken from the BayesX manual, for details see
`BayesX`.

```{r Zambia, fig.height=8, fig.width=8, dev='png'}
data(Zambia)

#load map
map <- suppressWarnings(read.bnd("Zambia/zambia.bnd"))

plotMap = function(map, x, nbreaks = 5, theme = "Blues", ...){
  pal  <- brewer.pal(n = nbreaks, name = theme)
  brks <- quantile(x, probs = seq(0, 1, 1 / nbreaks))
  cols <- cut(x, breaks = brks, labels = pal)
  plot(map, col = as.character(cols), ...)
}

# add one column for the unstructured spatial effect
Zambia$distr.unstruct  =  Zambia$district

formula <- hazstd ~ f(inla.group(bmi), model = "rw2") + f(agc, model = "rw2") +
                        f(district, model = "besag", graph = "Zambia/zambia.graph", param = c(1, 0.01)) +
                        f(distr.unstruct, model = "iid", param = c(1, 0.01)) +
                        rcw + edu1 + edu2 + tpr + sex

mod.zambia <- inla(formula, data = Zambia, control.fixed = list(prec.intercept = 0.001, prec = 0.001), verbose=TRUE)

#map posterior means
posterior.means <- mod.zambia$summary.random$district$mean
plotMap(map, posterior.means)
title(main = "Posterior means")
```


