---
title: "magrittr"
author: "Rob Weyant"
date: "robert.weyant@gmail.com"
output: 
  ioslides_presentation:
    widescreen: true
    css: magrittr_custom.css
---


## Overview

### {magittr} provides 4 special operators

- `%>%` - pipe operators
- `%T>%` - tee operator
- `%$%` - exposition operator
- `%<>%` - compound assignment pipe operator
```{r, echo=FALSE, warning=FALSE,message=FALSE}
library(magrittr,quietly=T)
library(dplyr,quietly=T)
```

## Other Special Operators in R

- `%*%` - matrix multiplication, `x %*% y`
- `%in%` - value matching
- `%%` - modulus operator
- `%o%` - outer product, `outer()`
- `%x%` - Kronecker product, `kronecker()`
- `%/%` - integer division

## The Problem
#### R code can get hard to read
  
<br>
```{r, eval=FALSE}
### <b>
sapply(iris[iris$Sepal.Length < mean(iris$Sepal.Length),-5],FUN = mean)
### </b>  
```


## A (Possible) Solution - the pipe %>% {.build} 

- Similar to Unix pipe <b>|</b>
- Code can be written in the order of execution, left to right
- <b>`%>%`</b> will "pipe" information from one statement to the next  
  * <b>`x %>% f`</b> is equivalent to <b>`f(x)`</b>  
  * <b>`x %>% f(y)`</b> is equivalent to <b>`f(x,y)`</b>  
  * <b>`x %>% f %>% g %>% h`</b> is equivalent to <b>`h(g(f(x)))`</b>  

## What %>% is doing

The <b>`%>%`</b> is taking the output of the left-hand side and using that for
the first argument of the right-hand side, or where it finds a *`.`*  

Example using `head(x, ...)`

```{r}
mtcars %>% head(.,2)  # same as using head(mtcars,2)
mtcars %>% head(2)     # same as using head(mtcars,2)
```

## Slightly more complicated example

```{r,fig.height=3.5,fig.width=5,fig.align='center'}
library(ggplot2)
mtcars %>%
  xtabs(~gear+carb,data=.) %>% 
  as.data.frame %>% 
  ggplot(.,aes(x=gear,y=carb,size=Freq)) +
  geom_point()
```

## Even more complicated example

```{r, eval=FALSE}
# Generate some sample data.
df <-
    data.frame(
        Price    = 1:100 %>% sample(replace = TRUE),
        Quantity = 1:10  %>% sample(replace = TRUE),
        Type     =
            0:1 %>%
            sample(replace = TRUE) %>%
            factor(labels = c("Buy", "Sell"))
    ) 
```
[Source](http://www.r-statistics.com/2014/08/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/)


## The combination of %>% with {dplyr}

 - `filter()`
 - `summarize()`
 - `arrange()`
 - `mutate()`


## The combination of %>% with {dplyr}
```{r}
sapply(iris[iris$Sepal.Length < mean(iris$Sepal.Length),-5],FUN = mean)
iris %>% 
  mutate(avg.length=mean(Sepal.Length)) %>% 
  filter(Sepal.Length<avg.length) %>% 
  select(-Species,-avg.length) %>%
  summarise_each(funs(mean))
```

## %$% The exposition operator

- Similar to `with()` or `attach()`  
- Useful for functions that don't take a data parameter
- Can execute several statements by wrapping them in `{}`

## %$% The exposition operator
```{r}
library(datasets)
table(CO2$Treatment,CO2$Type)
# with(CO2,table(Treatment,Type))
CO2 %$% table(Treatment,Type)
```

## %T>% The Tee Operator

- Allows a "break" in the pipe.  
- Executes right-hand side of `%T>%`, but will continue to pipe through to next statement

## %T>% The Tee Operator

```{r, fig.align='center',fig.height=2.7,fig.width=5}
iris %>%
  filter(Species != 'virginica') %>% 
  select(Sepal.Width,Sepal.Length) %T>%
  plot %>%  # Make scatterplot and keep going
  colMeans
```

## %<>% The Compound Assignment Operator
- Combines a pipe and an assignment operator
- Think <b>`i++`</b> or <b>`x+=z`</b> from the C family, Python, Ruby, etc.

## %<>% The Compound Assignment Operator
```{r}
df <- rexp(5,.5) %>% data.frame(col1=.)
df
df %<>% arrange(col1)
df
```

## Other things to be aware of

- Shortcut for `%>%`: Ctrl+Shift+m
- `?%>%` does not work, use `?'%>%'`
- `df %>% .something. %>% sytem.time` does not work

## Links

- [magrittr on GitHub](https://github.com/smbache/magrittr)  
- [dplyr and pipes: the basics](http://seananderson.ca/2014/09/13/dplyr-intro.html)
- [pipeR vs. magrittr](http://renkun.me/blog/2014/08/08/difference-between-magrittr-and-pipeR.html)



# Thank You