---
title: "Package Development and spotifyr"
author: "Rob Weyant"
date: "11/9/2017"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Introduction

## Why develop packages?

* Easier to share, easy to use

* Familiar/standard structure 

* Easy to reuse code between projects

  * Very practical within companies, research groups, collaborators 
  
  * Do it once, correctly.
  
---

# So you want to make a package

## Step 1: Think of a sweet name

* R puns encouraged

* otherwise, see [Hadley's recommendations](http://r-pkgs.had.co.nz/package.html)

  * googleable (if public)
  
  * indicate its use

## Step 2: Think of what you want to do

* Maybe this is the first step

## Step 3: Start coding

---

# Create Your Package (devtools)

```{r, eval=FALSE}
library(devtools)
create('coolpackage')
```

Barebones package structure:

```{bash}
ls ~/coolpackage
```

---

# Create Your Package (RStudio)

## File --> New Project --> New Directory --> R Package

Slightly less barebones package structure:

```{bash}
ls ~/coolpackage2
```

---

# Package Components

* DESCRIPTION: metadata about the project

* NAMESPACE: directives that control the namespace of the project

* R/: Where the R code goes

* man/: where the documentation goes

* tests/: where tests go

---

# spotifyr

Wrapper for Spotify's [Web API](https://developer.spotify.com/web-api/)

... in R!

## But... why?

* learning experience

* Spotify has lots of data!

* New beta features let you control spotify!

* not first package I worked on, but first one I cared about

---

# DESCRIPTION

#### `Imports`

* Need to be installed for your package to function

* Packages in the `Imports` section get loaded when the package is attached

#### `Suggests`

* Nice to have, but not required.

* Why use this?

---

# R/

* Any code in this directory will be executed and saved when the package is built.

  * Ctrl+Shift+B

* Any code in this directory will be _sourced_ `devtools::load_all()`. That is, executed and the results made available

  * Ctrl+Shift+L

---

# Documentation

### [roxygen!](http://roxygen.org/)

* Add in-source documentation that gets translated to Rd files

* devtools::document()

* *Generate Documentation*: Ctrl+Shift+D / Cmd+Shift+D

* Will generate and update NAMESPACE file

* Can execute examples in comments!

---

# tests/

### [testthat!](http://r-pkgs.had.co.nz/tests.html)

* devtools::use_testthat()

* add test files into `./tests/testthat/`
  
  * filename: `test*`

* `expect_*`
  
  * `expect_equals`, `expect_gt`, `expect_is`

* *Run all tests*: Ctrl+Shift+T / Cmd+Shift+T

---

# Resources

* http://r-pkgs.had.co.nz/

* https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/


