---
title: "Spotifyr Project"
author: "Rob Weyant"
output: 
  ioslides_presentation:
    widescreen: true
    css: custom.css
---


```{r, echo=FALSE,message=FALSE}
## Load package for presentation.  
library(spotifyr)
library(plyr)
options(width=120)
```

## About Me {.bigger}

- Data Scientist at Trinity Health

    + Support clinical research projects
    
    + Support new data warehousing projects
    
- Education: 

    + Math, Statistics (Penn State)
    
    + Biostatistics (UM)

- Raised in Pittsburgh, PA

- Interests: Skiing, R, Beer, Music

## [Spotify API](https://developer.spotify.com/web-api/endpoint-reference/) {.normal}

- HTTP-based RESTful API, with standard HTTP-Methods (GET, PUT, POST, DELETE)

- Get metadata on artists, albums, songs

- Allows interactions with Spotify users

    + Follow and unfollow artists and users

- Create and manage playlists

- Returns **JSON**

```{r fig.width=5,fig.height=2,fig.align='right', echo=FALSE}
suppressWarnings(suppressPackageStartupMessages(library(png)))
suppressPackageStartupMessages(library(grid))
img <- readPNG('figures/spotify_logo.png')
grid.raster(img)
```

## Basic Unauthenticated API Call {.smaller .blackcode }

**Install**
```{r, eval=FALSE}
options(unzip='internal') # If on Ubuntu 14.04+, openSUSE, ...
devtools::install_github('rweyant/spotifyr')
library(spotifyr)
```

Using Hadley's `httr` package:

- Form URL in specific pattern (**GET**): `https://api.spotify.com/v1/artists/{artist id}`

  e.g. [https://api.spotify.com/v1/artists/4mTFQE6aiehScgvreB9llC](https://api.spotify.com/v1/artists/4mTFQE6aiehScgvreB9llC)

```{r,comment=''}
# Search Spotify's data for artists with names like "earth" and play drone (GET)
search_artist <- spotifyr::search('earth genre:drone',type = 'artist')

# Helper function to extract JSON elements into data.frame
summarized_artists <- spotifyr::simplify_result(search_artist$artists, type='artists')
summarized_artists[,c('name','id','genre','popularity','followers')]
```

## More data {.smaller .blackcode}

```{r,comment=''}
top_tracks <- spotifyr::get_artist_toptracks(summarized_artists$id[1],'us')
summarized_tracks <- spotifyr::simplify_result(top_tracks$tracks,type='songs')
summarized_tracks[1:2,c('artists.name','name','popularity','duration_ms','album.name')]
```
<iframe src="https://embed.spotify.com/?uri=spotify:track:7kRY2GFOWywaWFE1drt80E" width="250" height="120" frameborder="0" allowtransparency="true"></iframe>


## Future Plans {.bigger}

- Test and improve stability across different operating systems

- Work on the supporting the last 2 API endpoints for playlist management that still don't work 

- Improve helper functions to simplify returned JSON

- Create a music recommender Shiny app integrating [Gracenote](https://developer.gracenote.com/music-platforms), [last.fm](http://www.last.fm), and [metal-archives](metal-archives.com) data with embedded Spotify player and playlist creater 

[https://github.com/rweyant/spotifyr](https://github.com/rweyant/spotifyr)



