---
title: "CSS example"
author: "AARUG"
date: "March 29, 2016"
output: 
  ioslides_presentation:
    css: aarug.css
---

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring **HTML**, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Smaller Text {.smaller}

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.


## Unbolded Text {.unbolded}

This is an R **Markdown** presentation. Markdown is a simple formatting syntax for authoring **HTML**, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any **embedded R code** chunks within the document.


## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with Numbered Bullets

1. Bullet 1
2. Bullet 2
3. Bullet 3

## Slide with multiple classes {.huge .hugebullets}

Add multiple classes within curly brackets like {.huge .hugebullets}

- Huge Bullet 1

- Huge Bullet 2

- Huge Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```

## Slide with smaller code {.smallercode}

```{r, messages = FALSE} 
suppressPackageStartupMessages(library(dplyr))
tbl_df(cars) %>% filter(speed > -10) 
```


## Slide with smallest code {.smallestcode}

```{r, messages = FALSE} 
suppressPackageStartupMessages(library(dplyr))
tbl_df(cars) %>% filter(speed > -10) 
```

